```{r}
#| label: setup
#| include: false
library(here)
source(here("sections", "setup.R"))
```

::: {.cols-2}

Before the company re-branded in 2022, Posit (then RStudio) was often
thought of as an "R company" in the data community because of its
dedication to developing and maintaining some of the most used R
libraries in the world. However, Posit has always been better described
as a *scientific software* company. Supporting Python (via the
[reticulate](https://github.com/rstudio/reticulate)
package, RStudio language support), working with relational databases
and data platforms such as Apache Spark (a cross-platform data frame
compatibility via [feather/Apache
Arrow](https://github.com/apache/arrow)), and much more
mean that we've never been solely an "R company".

{{< colbreak >}}

Since the 2023 PBC report, Posit has released several new packages in
the Python and R ecosystems and continues to maintain and grow the
libraries previously developed. The following subsections highlight
selected Posit software projects of interest to the broad data science
community. Where metrics are published, please note these represent a
*lower bound* on the actual number, as it is difficult-to-impossible to
account for every install and usage in the world.

:::

```{=typst}
#place(bottom, dy: 1.25in,
  align(center, 
    image("assets/images/2-0616.jpg", 
      width: 100% + 2.5in, fit: "cover")
  )
)
```

{{< pagebreak >}}

::: {.page-banner image_paths='("assets/images/quarto-logo-dark.svg",)' image_height="50%"}

## Quarto

::: {.cols-2}

In July 2022, [Posit
announced](https://posit.co/blog/announcing-quarto-a-new-scientific-and-technical-publishing-system/)
the [Quarto](https://quarto.org/) project, an open-source
scientific and technical publishing system as a successor to the [R
Markdown](https://rmarkdown.rstudio.com/) library. While
Quarto incorporates the lessons learned from over 10 years of developing
R Markdown into an entirely new project, it's likely still quite
familiar to users of R Markdown as they share two core dependencies:
Knitr and Pandoc. In fact, despite the fact that Quarto does some things
differently, most existing R Markdown documents can be rendered
unmodified using Quarto.

{{< colbreak >}}

Quarto allows users to choose from multiple computational engines
(Knitr, Jupyter, and Observable), which makes it easy to use Quarto with
R, Python, Julia, JavaScript and many other languages. It also allows
users to author documents as plain text markdown or Jupyter Notebooks,
and publish to numerous outputs such as HTML, PDF, MS Word, ePub and
more. Finally, the community has already extended Quarto , as shown by
the
[quartodoc](https://machow.github.io/quartodoc/get-started/overview.html)
project for developing API documentation.

There are 4 full time equivalent (FTE) employees developing open-source
Quarto products as of May 2025.

:::
 
```{r}
#| label: quarto
quarto |> 
  mutate(major_minor = if_else(major_minor == 1.8, "1.8\nPre-release", as.character(major_minor))) |> 
  ggplot(aes(major_minor, cumsum(downloads))) +
  geom_line(aes(group = 1), color = posit_colors$blue, linewidth = 0.75) +  
  plot_defaults
```

::: 

::: {.page-banner image_paths='("assets/images/shiny-logo.png",)' image_height="50%"}

## Shiny

::: {.cols-2}

Shiny has been a mainstay in the R community since its launch in 2012,
providing a web application framework that makes it easy to tell data
stories in interactive point-and-click web applications. In April 2023,
Posit released the Python version of Shiny, bringing the same great
reactive programming model and modular design to the PyData ecosystem.  
More recently, the [Shiny Express](https://shiny.posit.co/py/docs/express-in-depth.html) 
syntax was introduced. It offers a steamlined syntax that makes it easier 
for Python developers to get started with Shiny.

New packages (listed in a later section) also highlight Shiny as a strong 
platform for building chat-based and other LLM-powered apps.

{{< colbreak >}}

Shiny applications can be shared with others via an open-source [Shiny
Server](https://posit.co/products/open-source/shinyserver/),
the hosted [shinyapps.io](http://shinyapps.io) service,
with [Posit Connect Cloud](https://connect.posit.cloud/) or [Posit
Connect](https://posit.co/products/enterprise/connect/).
Shiny and related packages include shiny
([Python](https://shiny.posit.co/py/),
[R](https://shiny.posit.co/r/getstarted)),
[bslib](https://rstudio.github.io/bslib/),
[shinytest](https://rstudio.github.io/shinytest/),
[shinyloadtest](https://rstudio.github.io/shinyloadtest/),
[shinydashboard](https://rstudio.github.io/shinydashboard/),
[leaflet](https://rstudio.github.io/leaflet/), and
[crosstalk](https://rstudio.github.io/crosstalk/).

There are 5 FTE Posit employees developing the open-source Shiny and
Shiny Server products as of May 2025.

:::

```{r}
#| label: shiny
#| fig-height: 2.75
python |> 
  filter(package == "shiny") |> 
  growth_plot() + ggtitle("Shiny for Python")
r |> filter(package == "shiny") |> growth_plot() + ggtitle("Shiny for R")
```

::: 

{{< pagebreak >}}

::: {.page-banner image_paths='("assets/images/hexes/ellmer.svg","assets/images/hexes/ragnar.png", "assets/images/hexes/ragnar.png", "assets/images/hexes/gander(lores).png", "assets/images/hexes/mall.png")'}

## AI and LLM Enablement

::: {.cols-2}

LLMs (large language models) are changing how data scientists work. 
Posit's Open Source teams are building tools to help data scientists responsibly
use LLMs in their analysis, leveraging them to help during development, 
and incorporating LLM capabilities in the solutions they provide others.

### Selection of packages for enabling LLMs in data science applications

- [Ellmer](https://ellmer.tidyverse.org/) makes it easy to use large language models (LLM) from R. It supports a  variety of LLM providers and implements a rich set of features including streaming outputs, tool/function calling and structured data extraction.
- [Chatlas](https://posit.co/blog/announcing-chatlas/) is a flexible Python interface to many LLM providers (playing a similar role to ellmer). It supports tool use, function calling, and streaming responses.
- [Ragnar](https://ragnar.tidyverse.org) brings Retrieval-Augmented Generation (RAG) to R. Helps users index their own data and get LLM responses with grounded answers.
- [Querychat](https://github.com/posit-dev/querychat) adds an SQL-powered LLM to Shiny apps. It lets users explore data with natural language. Querychat works in both R and Python.

### Selection of packages that assist during development

- [ShinyAssistant](https://shiny.posit.co/blog/posts/shiny-assistant/) helps prototype Shiny apps using a simple chat interface leveraging LLMs to generate entire applications.
- [Gander](https://posit.co/blog/introducing-gander/) is a coding assistant that understands R environments and shares context like column names and types to improve help quality.
- [Chores](https://posit.co/blog/introducing-chores/) connects Ellmer to your source editor in RStudio and Positron. I automates repetitive programming tasks.
- [Mall](https://posit.co/blog/mall-ai-powered-text-analysis/) enables LLM powered sentiment analysis, text summarization, text classification, information extraction and text translation.   Mall is available for both R and Python.

LLMs are also integrated as coding assistants into RStudio and Positron.
Positron is Posit's new IDE described below.

As of May 2025, 4 FTE Posit employees are working  
on open-source tools related to LLMs.

:::

::: {.page-banner image_paths='("assets/images/hexes/gt.svg",)'}

## gt / Great Tables

::: {.cols-2}

When presenting an analysis, a table can often convey the results more
concisely than the most beautiful and interactive of charts. However,
the experience of creating and displaying tables in R and Python has
been mixed, especially when you want to display something beyond a plain
data frame representation.

{{< colbreak >}}

To that end, the [gt](https://gt.rstudio.com/) and [Great
Tables](https://posit-dev.github.io/great-tables/articles/intro.html)
packages have defined a "grammar of tables" to solve this problem (in R
and Python, respectively), analogous to the "grammar of graphics" for
specifying charts.

As of May 2025, there is 1 FTE Posit employee developing gt / Great
Tables open-source packages.

:::

```{r}
#| label: gt
python |> filter(package == "great-tables") |> growth_plot() + ggtitle("great_tables")
r |> filter(package == "gt") |> growth_plot() + ggtitle("gt")
```

:::

::: {.page-banner image_paths='("assets/images/hexes/pins.svg",)'}

::: {.page-banner image_paths='("assets/images/hexes/plotnine.png",)'}

## Plotnine

::: {.cols-2}

[Plotnine](https://plotnine.org/) is an implementation of
the grammar of graphics in Python, heavily influenced by ggplot2 in R.
Built upon the ubiquitous
[matplotlib](https://matplotlib.org/) plotting library,
custom (and otherwise complex) plots are easy to reason about and build
incrementally, while the simple plots remain simple to create.

There is 1 FTE Posit employee developing Plotnine as of May 2025.

:::

```{r}
#| label: plotnine
python |> filter(package == "plotnine") |> growth_plot() + ggtitle("Plotnine")
```

:::

## Pins

::: {.cols-2}

Pins (for [R](https://pins.rstudio.com/) and [python](https://rstudio.github.io/pins-python/))
publish data, models, and other objects, making them easy to share across projects and with other. 
Users can pin objects to a variety of pin boards, including folders (to share on a networked 
drive or with services like DropBox), Posit Connect, Amazon S3, and Google Cloud Storage. 
Pins can be automatically versioned, making it straightforward to track changes, 
re-run analyses on historical data, and undo mistakes.

::: 

```{r}
#| label: pins
python |> filter(package == "pins") |> growth_plot() + ggtitle("pins for Python")
r |> filter(package == "pins", year(date) > 2021) |> growth_plot() + ggtitle("pins for R")
```

:::

::: {.page-banner image_paths='("assets/images/hexes/vetiver.svg",)'}

## Vetiver

::: {.cols-2}

[Vetiver](https://vetiver.posit.co/) solves the issues
around versioning, sharing, deploying and monitoring predictive models
served via APIs. Available for both R and Python, vetiver is extensible
via generics that support many common types of models. {{< colbreak >}} Vetiver also
provides the "model cards" functionality, which can help to generate
documentation by extracting information about the generated model.

::: 

```{r}
#| label: vetiver
python |> filter(package == "vetiver") |> growth_plot() + ggtitle("vetiver for Python")
r |> filter(package == "vetiver", year(date) > 2021) |> growth_plot() + ggtitle("vetiver for R")
```

:::

```{=typst}
#v(6em)
#place(right, dy: -3em, image("assets/images/hexes/webr.svg", height: 5em))
```

:::

::: {.page-banner image_paths='("assets/images/hexes/plotnine.png",)'}

## Orbital

::: {.cols-2}

[Orbital](https://github.com/posit-dev/orbital) lets you run machine learning 
models inside your database.  Originally an [R package](https://orbital.tidymodels.org/articles/orbital.html), 
it now also supports Python.  In Python, orbital converts scikit-learn models into 
SQL, so they can run directly in a database like Snowflake—no Python environment needed.  
The performance gains this approach has provided is quite signficant.    

:::

## webR

::: {.cols-2}

[WebR](https://docs.r-wasm.org/webr/latest/) has the
ambitious goal of bringing the R language to the browser, removing the
need for a backend server for computation. It also allows for
computation to be done on the client machine, supporting use cases that
are infeasible or undesirable for using server-side processing (such as
not wanting to send personal data over the internet). {{< colbreak  >}}
Also, by making
the most of the user's device capabilities, webR can improve performance
and lower app hosting costs.

There is 1 FTE Posit employee developing enterprise focused open-source 
products like orbital as of May 2025.

:::

:::



::: {.page-banner image_paths='("assets/images/hexes/tidyverse.svg", )'}

## Tidyverse

::: {.cols-2}

The [tidyverse](https://www.tidyverse.org/) is an
opinionated collection of R packages designed for data science. All
packages share an underlying design philosophy, grammar and data
structures. 

The tidyverse consists of nine core packages (including ggplot2, tidyr
and readr) and 31 packages overall.

{{< colbreak >}}

There are 9 FTE Posit employees developing Tidyverse and related
open-source products as of May 2024.

:::

```{r}
#| label: tidyverse
r |> 
  filter(project == "tidyverse") |> 
  group_by(date) |> 
  summarise(downloads = sum(downloads)) |> 
  growth_plot() + 
  ggtitle("Tidyverse")
```

:::

::: {.page-banner image_paths='("assets/images/hexes/tidymodels.svg", )'}

## Tidymodels

::: {.cols-2}

[Tidymodels](https://www.tidymodels.org/) is a cohesive
collection of packages that perform tasks relevant to statistical
modeling and machine learning. Tidymodels packages share a common syntax
and design philosophy, and are designed to work seamlessly with
Tidyverse packages.

{{< colbreak >}}

There are currently 42 tidymodels packages on CRAN. Popular tidymodels
packages include parsnip, rsample, recipes, tune and yardstick.

There are 3 FTE Posit employees developing Tidymodels and related
open-source products as of May 2025.

:::

```{r}
#| label: tidymodels
r |> 
  filter(project == "tidymodels") |> 
  group_by(date) |> 
  summarise(downloads = sum(downloads)) |> 
  growth_plot() + 
  ggtitle("Tidymodels")
```

:::

::: {.page-banner image_paths='("assets/images/hexes/sparklyr.svg", "assets/images/hexes/reticulate.svg", "assets/images/hexes/tensorflow.svg")'}

## Connectivity Packages 

::: {.cols-2}

Posit increases the efficiency of customers by making open-source
packages that connect data scientists to spreadsheets, databases,
distributed storage frameworks for big data, machine learning platforms,
and the programming environments of other languages, like python.

{{< colbreak >}}

Connectivity packages include:
[sparklyr](https://spark.posit.co/), [tensorflow for R](https://tensorflow.rstudio.com/),
[keras](https://keras.posit.co/),
[odbc](https://solutions.posit.co/connections/db/r-packages/odbc/),
and [reticulate](https://rstudio.github.io/reticulate/).

There are 3 FTE Posit employees creating connectivity-related
open-source packages as of May 2025.

:::

```{r}
#| label: connectivity
r |> 
  filter(project == "connectivity") |> 
  group_by(date) |> 
  summarise(downloads = sum(downloads)) |> 
  growth_plot() + 
  ggtitle("R Connectivity")
```

:::

::: {.page-banner image_paths='("assets/images/hexes/devtools.svg","assets/images/hexes/usethis.svg", "assets/images/hexes/roxygen2.svg", "assets/images/hexes/testthat.svg", "assets/images/hexes/pkgdown.svg")'}

## R Infrastructure Tools (r-lib)

::: {.cols-2}

R-lib is a large collection of R packages that make it easier to build,
find, and use effective tools for data analysis.

```{r}
#| include: false
#| eval: false
repos_json <- gh::gh("/orgs/{org}/repos", org = "r-lib", .limit = Inf)
names <- sapply(repos_json, "[[", "name")
length(intersect(names, rownames(available.packages())))
```

{{< colbreak >}}

There are currently 114 R-lib packages. Popular packages
include [devtools](https://devtools.r-lib.org/),
[testthat](https://testthat.r-lib.org/),
[roxygen2](https://roxygen2.r-lib.org/),
[pkgdown](https://pkgdown.r-lib.org/) and
[usethis](https://usethis.r-lib.org/).

:::

```{r}
#| label: rlibs
r |> 
  filter(project == "r-lib") |> 
  group_by(date) |> 
  summarise(downloads = sum(downloads)) |> 
  growth_plot() + 
  ggtitle("R Libs")
```

::: 

::: {.page-banner image_paths='("assets/images/hexes/RStudio.svg",)'}

## RStudio Integrated Development Environment

::: {.cols-2}

[RStudio](https://posit.co/products/open-source/rstudio/)
is a multi-language IDE designed for Data Science with R and Python. It
augments the standard code console with an editor that can display
Notebooks, launch apps, highlight code syntax, spot code errors, and
directly execute code. Built into the IDE are tools for debugging,
plotting, browsing files, and managing project histories and workspaces.
Together these tools make data scientists and developers much more
efficient.

{{< colbreak >}} 

There are 5 FTE Posit employees developing the RStudio IDE open-source
desktop and server products as of May 2024.

:::

```{r}
#| label: rstudio
rstudio |> growth_plot() + ggtitle("RStudio")
```

:::

# Additional Posit Free-to-Use Solutions

::: {.page-banner image_paths='("assets/images/hexes/positron.png",)'}

## Positron

::: {.cols-2}

[Positron](https://positron.posit.co//)
is a new multi-language IDE designed for Data Science.  Positron has first-class, built-in 
support for R and/or Python via an integrated console, with extensibility options for other 
languages. This native support includes specialized views and panes throughout 
Positron such as a [Data Explorer](https://positron.posit.co/data-explorer.html), 
[Connections Pane](https://positron.posit.co/connections-pane.html), Variables Pane, 
access to AI/LLM driven data assistants and more.

Positron separates the language interpreter from the IDE itself, which makes it more robust 
during development—if R or Python encounters an error, the IDE remains unaffected. This
architecture also allows a user to switch between different versions of their preferred language 
without needing to reload the entire IDE.

Positron is built on [Code OSS](https://github.com/microsoft/vscode) and supports VS Code 
compatible extensions (.vsix files), providing extensibility of capabilities beyond the core 
IDE itself. By building on Code OSS, Positron gets rich text editor capabilities and access 
to 1,000s of community extensions out of the box. 

Additional languages typically used in package development are supported via existing third party 
extensions. These don’t make use of the full Positron data science experience including an interactive 
console, plots, and similar. Some examples include Rust, Javascript/Typescript, C/C++, or Lua.

Positron™ is licensed under the [Elastic License 2.0](https://github.com/posit-dev/positron?tab=License-1-ov-file#readme), 
a source-available license. [Read more](https://positron.posit.co/licensing.html) about what 
this license means and our decision to use it.

Positron is deeply focused on native data science workflows; it provides a batteries-included 
and cohesive experience beyond that of a general-purpose IDE or text editor such as VS Code.

A beta version is currently available for Windows and macOS, with a full release planned for August 2025.

{{< colbreak >}} 

There are 14 FTE Posit employees developing Positron as of May 2025.

:::

:::

## Posit Public Package Manager

::: {.cols-2}

WIth the ubiquity of open source software in our daily lives, one area
that most people don't think about is 'How do you distribute that
software quickly and securely to the end user?'. To that end, Posit
created [Posit Package
Manager](https://posit.co/products/enterprise/package-manager/),
which gives companies a means for providing curated repositories,
repository snapshots for better reproducibility, the ability to air-gap
the repository for enhanced security and much more.

{{< colbreak >}}

As part of our commitment to improving the quality and availability of
open source software for all, Posit hosts a public instance of Posit
Package Manager called [Posit Public Package
Manager](https://packagemanager.posit.co/client/#/) that
mirrors CRAN, PyPI and Bioconductor. This mirror served nearly 40
million downloads per month in Q1 2024.

:::



